\paperTitle

### 2014-05-17 doc/icml2014/background.tex
Updated bottlenecked definition
terninology change
finished going through recovering hidden marginals

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\paragraph{Approach overview}
%
%\algorithmref{overview} summarizes our approach.
%
%\begin{algorithm}
%  \caption{\LearnMarginals}
%  \label{algo:overview}
%  \begin{algorithmic}
%    \REQUIRE Graphical model $\sG$, data $\sD$
%    \ENSURE Parameters $\theta$
%    \For{each clique $\sN$}
%    \FOR{each hidden node $h_i \in S$ with exclusive view $x_v$} 
%      \STATE Apply $\TensorFactorize$ to learn conditional moments
%      $\mOpp{v}{i} = \Pr(x_v \mid h_i)$.
%    \ENDFOR
%    \STATE Apply $\LearnClique$ to learn the marginals $\mH_\sC$.
%  \end{algorithmic}
%\end{algorithm}

@@

\begin{definition}[Bottleneck]
  \label{def:bottleneck}
  A hidden variable $h_i$ is said to be a \emph{bottleneck} if (i) there
  exists three observed variables (views), $x_{v_1}, x_{v_2}, x_{v_3}$,
  that are conditionally independent given $h_i$,
  and (ii) each $\mOpp{v}{i} \eqdef \Pr(x_v \mid h_i) \in \Re^{d \times k}$ has full column rank $k$
  for each $v \in \{v_1, v_2, v_3\}$.
  %Let $\sV_{h_i}$ denote any set of views for $h_i$.
  We say that a subset of hidden variables $S \subseteq \bh$ is a bottleneck
  if every $h \in S$ is a bottleneck.
  We say that a graphical model $\sG$ is bottlenecked if all its hidden variables
  are bottlenecks.
\end{definition}
For example, in \figureref{approach}, $x_1^a,x_1^b,x_2^a$ are views of the bottleneck $h_1$,
and $x_2^a,x_2^b,x_1^b$ are views of the bottleneck $h_2$.  Therefore, the clique
$\{h_1,h_2\}$ is a bottleneck.  Note that views are allowed to overlap.

@@

To simplify notation, consider the example in \figureref{three-view} where $h_1=1,v_1=1,v_2=2,v_3=3$.
The observed moments $\mO_{12}, \mO_{23}, \mO_{13}$ and $\mO_{123}$ can be factorized as follows:
\begin{align*}
  \mO_{vv'} &= \sum_h {\pi \oft 1}[h] {\mOppt{v}{1}}[h] \otimes {\mOppt{v'}{1}}[h] \\ %\quad \forall i,j \in \{1,2,3\} \\
  \mO_{123} &= \sum_h {\pi \oft 1}[h] {\mOppt{1}{1}}[h] \otimes {\mOppt{2}{1}}[h] \otimes  {\mOppt{3}{1}}[h].
\end{align*}
The $\TensorFactorize$ algorithm first computes a whitening matrix $W \in
  \Re^{d \times k}$ such that $W^\top \mO_{12} W = I_{k \times k}$,
  and uses $W$ to transform $\mO_{123}$ into a symmetric orthogonal tensor.  
Then a robust tensor power method is used to extract the eigenvectors
of the whitened $\mO_{123}$; unwhitening yields the columns of $\mOpp{3}{1}$
(up to permutation and scaling).
The other conditional moments can be recovered similarly.
%  we call $\TensorFactorize$, which transforms estimates of the observed moments 
%  into estimates of the parameters, $\pi\oft 1, \mOpp{1}{1},
%  \mOpp{2}{1},\mOpp{3}{1}$ (up to permutation of the columns), provided
%  the following assumption holds:

%It is easy to see that $\LearnMarginals$ has a polynomial sample complexity
%because it composes two parts that individually have polynomial sample
%complexity.
%From \citet{anandkumar13tensor} given the non-degeneracy assumptions
  %\assumptionref{non-degeneracy},
  The resulting estimate of %$\mOpphat{v}{i}$
  $\mOpp{v}{i}$ based on $n$ data points converges at a rate of $n^{-\frac12}$ with a constant
  that depends polynomially on $\sigma_k(\mOpp{v}{i})^{-1}$,
  the inverse of the $k$-th largest singular value of $\mOpp{v}{i}$.
% PL: we don't have a crisp theorem (because there's multiple paths), so leave it vague.
%Recovering the marginals $Z_\sC$ from the conditional moments via $\LearnClique$ is a linear operation and thus also has polynomial sample complexity. 
Note that $\sigma_{k}(\mOpp{v}{i})$ can become quite
small if $h_i$ and $x_v$ are connected via many intermediate hidden
variables.\footnote{To see this, suppose $h_1$ has a view $x_v$ via a chain:
$h_1 - h_2 \cdots - h_t - x_v$. In this example, if
$\sigma_k(\Pr(h_{i+1} \given h_{i})) = a_k$ for each $i = 1,
\cdots, t-1$, then $\sigma_k(\mOpp{v}{1}) = a_k^t \sigma_k(\mOpp{v}{t})$.}

>>>>

@@

\begin{definition}[Bottleneck]
  \label{def:bottleneck}
  A hidden variable $h_i$ is said to be a \emph{bottleneck} if (i) there
  exists three observed variables (views), $x_{v_1}, x_{v_2}, x_{v_3}$,
  that are conditionally independent given $h_i$,
  and (ii) each $\mOpp{v}{i} \eqdef \Pr(x_v \mid h_i) \in \Re^{d \times k}$ has full column rank $k$
  for each $v \in \{v_1, v_2, v_3\}$.
  %Let $\sV_{h_i}$ denote any set of views for $h_i$.
  We say that a subset of hidden variables $S \subseteq \bh$ is bottlenecked
  if every $h \in S$ is a bottleneck.
  We say that a graphical model $\sG$ is bottlenecked if all its hidden variables
  are bottlenecks.
\end{definition}
For example, in \figureref{approach}, $x_1^a,x_1^b,x_2^a$ are views of the bottleneck $h_1$,
and $x_2^a,x_2^b,x_1^b$ are views of the bottleneck $h_2$.  Therefore, the clique
$\{h_1,h_2\}$ is bottlenecked.  Note that views are allowed to overlap.

@@

To simplify notation, consider the example in \figureref{three-view} where $h_1=1,v_1=1,v_2=2,v_3=3$.
The observed moments $\mO_{12}, \mO_{23}, \mO_{13}$ and $\mO_{123}$ can be factorized as follows:
\begin{align*}
  \mO_{vv'} &= \sum_h {\pi \oft 1}[h] {\mOppt{v}{1}}[h] \otimes {\mOppt{v'}{1}}[h] \\ %\quad \forall i,j \in \{1,2,3\} \\
  \mO_{123} &= \sum_h {\pi \oft 1}[h] {\mOppt{1}{1}}[h] \otimes {\mOppt{2}{1}}[h] \otimes  {\mOppt{3}{1}}[h].
\end{align*}
The $\TensorFactorize$ algorithm first computes a whitening matrix $W \in
  \Re^{d \times k}$ such that $W^\top \mO_{12} W = I_{k \times k}$,
  and uses $W$ to transform $\mO_{123}$ into a symmetric orthogonal tensor.  
Then a robust tensor power method is used to extract the eigenvectors
of the whitened $\mO_{123}$; unwhitening yields the columns of $\mOpp{3}{1}$
(up to permutation).
The other conditional moments can be recovered similarly.
%  we call $\TensorFactorize$, which transforms estimates of the observed moments 
%  into estimates of the parameters, $\pi\oft 1, \mOpp{1}{1},
%  \mOpp{2}{1},\mOpp{3}{1}$ (up to permutation of the columns), provided
%  the following assumption holds:

%It is easy to see that $\LearnMarginals$ has a polynomial sample complexity
%because it composes two parts that individually have polynomial sample
%complexity.
%From \citet{anandkumar13tensor} given the non-degeneracy assumptions
  %\assumptionref{non-degeneracy},
  The resulting estimate of %$\mOpphat{v}{i}$
  $\mOpp{v}{i}$ based on $n$ data points converges at a rate of $n^{-\frac12}$ with a constant
  that depends polynomially on $\sigma_k(\mOpp{v}{i})^{-1}$,
  the inverse of the $k$-th largest singular value of $\mOpp{v}{i}$.
% PL: we don't have a crisp theorem (because there's multiple paths), so leave it vague.
Note that $\sigma_{k}(\mOpp{v}{i})$ can become quite
small if $h_i$ and $x_v$ are connected via many intermediate hidden
variables.\footnote{To see this, suppose $h_1$ has a view $x_v$ via a chain:
$h_1 - h_2 \cdots - h_t - x_v$. In this example, if
$\sigma_k(\Pr(h_{i+1} \given h_{i})) = a_k$ for each $i = 1,
\cdots, t-1$, then $\sigma_k(\mOpp{v}{1}) = a_k^t \sigma_k(\mOpp{v}{t})$.}

### 2014-05-17 doc/icml2014/directed.tex
Fixed exclusive views
finished going through recovering hidden marginals

  \caption{(a) A bottleneck can be estimated consistently
  using tensor factorization \citep{anandkumar13tensor}.
  (b) An example of exclusive views for a bi-dependent subset $S$.
  }
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph{Structural properties}

We have shown how to estimate the parameters of any clique that possesses the exclusive
  views property (\definitionref{exclusive-views}).
  But in a general graph, which cliques have this property?
  To provide intuition, we will provide more interpretable sufficient conditions
  and examples.

% Define bidependent
First, define the notion of a bidependent set (in analogy with biconnected component),
in which conditionining on one variable does not break the set apart:
\begin{definition}[Bidependent set]
Given a graphical model $\sG$, we say that a subset of nodes $S$ is \emph{bidependent} if
conditioned on any $a \in S$, there is an active trail between any other two nodes $b,c \in S$.
\end{definition}
Note that all cliques are bidependent, but bidependent sets can have more conditional independences.
This will be important for pseudolikliehood in \sectionref{pseudolikelihood}.

% bidependent, bottleneck => exclusive views
Bidependent sets are significant because they guarantee exclusive views if they are bottlenecks:
%If a subset of hidden nodes $S$ is a bottleneck, it is clear that (ii)
%of the exclusive views property (\definitionref{exclusive-views}) via \TensorFactorize,
%but we will show that part (i) follows as well.
%Now we establish the link between bottlenecks and exclusive views:
\begin{lemma}[Bottlenecked implies exclusive views]
  \label{lem:bottleneck-views}  
  Let $S \subseteq \bh$ be a bidependent subset of hidden variables.
  If $S$ is a bottleneck, then $S$ has the exclusive views property.
\end{lemma}
\begin{proof}
Let $S$ be a bidependent subset and fix any $h_0 \in S$.
Since $h_0$ is a bottleneck, it has three conditionally independent views,
say $x_1, x_2, x_3$ without loss of generality. 
For condition (i), we will show that at least one of the views is conditionally independent
of $S \backslash \{ h_0 \}$ given $h_0$.
For the sake of contradiction, 
suppose that each observed variable $x_i$ is conditionally dependent on some
$h_i \in S \backslash \{h_0\}$ given $h_0$, for $i \in \{1, 2, 3\}$.
Then conditioned on $h_0$,
there is an active trail between $h_1$ and $h_2$ because $S$ is biconnected.
Thus, there is an active trail $x_1 - h_1 - h_2 - x_2$ conditioned on $h_0$.
Since the graphical model is faithful by assumption, we have $x_1 \not\perp x_2 \mid h_0$,
contradicting the fact that $x_1$ and $x_2$ are conditionally independent given $h_0$.
To show condition (ii), assume, without loss of generality, that $x_1$ is an exclusive view.
Then we can recover $\mOpp{1}{0} = \Pr(x_1 \mid h_0)$ via $\TensorFactorize$.
\end{proof}

@@

% DONE: move here because not central to story
\paragraph{Remarks.}
Note that even having two independent views for each $h_i \in S$ is sufficient for condition (i)
  of the exclusive views property, while three is needed for condition (ii).
The bottleneck property (\definitionref{bottleneck} can also be relaxed if some cliques
  share parameters.
For example, in the directed grid model, we can recover the conditional moments $O$ from
  $x^a_1, x^b_1$ and $x^a_2$ with $h_1$ as the bottleneck.
  Therefore, $h_2, h_3$ and $h_4$
  need not be bottlenecks, and we can omit the observations $x^b_2, x^b_3$ and $x^b_4$.

Our method also extends directly to case in which the observed variables
  are real-valued, while the hidden variables remain discrete. 
In this setting, the tensor factorization method of
  \citet{anandkumar13tensor} recovers the expected conditional means,
  $\mu_{vi} \eqdef \E(x_v | h_i) \in \Re^{d \times k}$ for each observed variable $x_v$ and
  hidden variable $h_i$.
\algorithmref{learnMarginals} % and \algorithmref{directed} 
still applies and allow us to recover clique marginals $Z_\sC$.
%  \todo{make this a footnote if running out of space}
% PL: don't understand this comment
%In this case, we only require that the hidden variables in distinct
  %cliques share parameters.

\begin{figure}
  \centering
  \subfigure[Hidden Markov model] {
    \includegraphics[width=0.45\columnwidth]{figures/hmm.pdf}
    \label{fig:examples-hmm}
  }
%  \subfigure[Directed grid model] {
%    \label{fig:examples-grid}
%    \includegraphics{figures/grid.pdf}
%  }
  \subfigure[Tree model] {
    \includegraphics[width=0.45\columnwidth]{figures/tree.pdf}
    \label{fig:examples-tree}
  }
  \subfigure[Noisy-or model] {
    \includegraphics[height=5em]{figures/non-example.pdf}
    \label{fig:examples-noisy-or}
  }
  \caption{(a) and (b): graphs that satisfy the exclusive views property; (c) does not.}
  \label{fig:examples}
\end{figure}

@@

In the HMM (\figureref{examples-hmm}), the parameters
are $O \eqdef \Pr(x_i|h_i)$ and $T \eqdef \Pr(h_{i+1} | h_i)$
for all $i$.
While the first and last hidden variables $h_1, h_M$ in the
  sequence are not bottlenecks, they still have exclusive views ($x_1$ and
  $x_M$, respectively)
  due to parameter sharing.

@@

In the latent tree model (\figureref{examples-tree}), the parameters
are $\pi \eqdef \Pr(h_i)$, $T \eqdef \Pr(h_i | h_1)$ for $i \in \{2,3,4\}$,
and $O \eqdef \Pr(x^a_i | h_i) = \Pr(x^b_i | h_i)$ for $i \in \{2,3,4\}$.
Note that while $h_1$ is not directly connected to an observed variable,
  it is still a bottleneck, with views $x^a_2, x^a_3, x^a_4$.
We can recover $T$ from the clique $\{h_1, h_2\}$ by using views $x^a_2$
  (exclusive to $h_2$) and $x^a_3$ (exclusive to $h_1$).

\paragraph{Non-examples}
\label{sec:non-example}

There are certainly models which are identifiable but do not have exclusive views.
For example, \figureref{examples-noisy-or} shows
  a binary-valued noisy-or network which can be
  learned by the algorithm of \citet{halpern13noisyor},
  but does not satisfy the exclusive views property.

>>>>

  \caption{
    \label{fig:bottleneckExclusiveViews}
  (a) A bottleneck can be estimated consistently
  using tensor factorization \citep{anandkumar13tensor}.
  (b) An example of exclusive views for a bi-dependent subset $S$.
  }
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph{Relationship to bottlenecks}

%We have shown how to estimate the parameters of any clique that possesses the exclusive
  %views property (\definitionref{exclusive-views}).
  %But in a general graph, which cliques have this property?
  %To provide intuition, we will provide more interpretable sufficient conditions
  %and examples.

The bottleneck property allow recovery of conditional moments,
and the exclusive views property allow recovery of hidden marginals.
But we will now show that the latter property is in fact implied by the former
property for special sets of hidden variables, which we call bidependent
sets (in analogy with biconnected components),
in which conditioning on one variable does not break the set apart:
\begin{definition}[Bidependent set]
We say that a subset of nodes $S$ is \emph{bidependent} if
conditioned on any $a \in S$, there is an active trail between any other two nodes $b,c \in S$.
\end{definition}
Note that all cliques are bidependent, but bidependent sets can have more conditional independences
(e.g., $\{h_1,h_2,h_3\}$ in \figureref{bottleneckExclusiveViews}(b)).
This will be important for pseudolikelihood in \sectionref{pseudolikelihood}.

% bidependent, bottleneck => exclusive views
Bidependent sets are significant because they guarantee exclusive views if they
are bottlenecked:
%If a subset of hidden nodes $S$ is a bottleneck, it is clear that (ii)
%of the exclusive views property (\definitionref{exclusive-views}) via \TensorFactorize,
%but we will show that part (i) follows as well.
%Now we establish the link between bottlenecks and exclusive views:
\begin{lemma}[Bottlenecked implies exclusive views]
  \label{lem:bottleneck-views}  
  Let $S \subseteq \bh$ be a bidependent subset of hidden variables.
  If $S$ is bottlenecked, then $S$ has the exclusive views property.
\end{lemma}
\begin{proof}
Let $S$ be a bidependent subset and fix any $h_0 \in S$.
Since $h_0$ is a bottleneck, it has three conditionally independent views,
say $x_1, x_2, x_3$ without loss of generality. 
For condition (i), we will show that at least one of the views is conditionally independent
of $S \backslash \{ h_0 \}$ given $h_0$.
For the sake of contradiction, 
suppose that each observed variable $x_i$ is conditionally dependent on some
$h_i \in S \backslash \{h_0\}$ given $h_0$, for $i \in \{1, 2, 3\}$.
Then conditioned on $h_0$,
there is an active trail between $h_1$ and $h_2$ because $S$ is biconnected.
This means there is also an active trail $x_1 - h_1 - h_2 - x_2$ conditioned on $h_0$.
Since the graphical model is faithful by assumption, we have $x_1 \not\perp x_2 \mid h_0$,
contradicting the fact that $x_1$ and $x_2$ are conditionally independent given $h_0$.
To show condition (ii), assume, without loss of generality, that $x_1$ is an exclusive view.
Then we can recover $\mOpp{1}{0} = \Pr(x_1 \mid h_0)$ via $\TensorFactorize$.
\end{proof}

@@

% DONE: move here because not central to story
\paragraph{Remarks.}
Note that having only two independent views for each $h_i \in S$ is sufficient for condition (i)
  of the exclusive views property, while three is needed for condition (ii).
The bottleneck property (\definitionref{bottleneck} can also be relaxed if some cliques
  share parameters (see examples below).
%For example, in the directed grid model, we can recover the conditional moments $O$ from
  %$x^a_1, x^b_1$ and $x^a_2$ with $h_1$ as the bottleneck.
  %Therefore, $h_2, h_3$ and $h_4$
  %need not be bottlenecks, and we can omit the observations $x^b_2, x^b_3$ and $x^b_4$.

Our method extends naturally to the case in which the observed variables
  are real-valued $x_v \in \Re^d$, as long as the hidden variables remain discrete. 
In this setting, the conditional moments
  $\mOpp{v}{i} \eqdef \E(x_v | h_i) \in \Re^{d \times k}$
  would no longer be distributions but general rank $k$ matrices.
  %for each observed variable $x_v$ and
  %hidden variable $h_i$.
%\algorithmref{learnMarginals} % and \algorithmref{directed} 
%still applies and allow us to recover clique marginals $Z_\sC$.
%  \todo{make this a footnote if running out of space}
% PL: don't understand this comment
%In this case, we only require that the hidden variables in distinct
  %cliques share parameters.

\begin{figure}
  \centering
  \subfigure[Hidden Markov model] {
    \includegraphics[width=0.45\columnwidth]{figures/hmm.pdf}
    \label{fig:examples-hmm}
  }
%  \subfigure[Directed grid model] {
%    \label{fig:examples-grid}
%    \includegraphics{figures/grid.pdf}
%  }
  \subfigure[Tree model] {
    \includegraphics[width=0.45\columnwidth]{figures/tree.pdf}
    \label{fig:examples-tree}
  }
  \subfigure[Noisy-or model] {
    \includegraphics[height=5em]{figures/non-example.pdf}
    \label{fig:examples-noisy-or}
  }
  \caption{(a) and (b): graphical models that satisfy the exclusive views property;
  (c) a graphical model that does not.}
  \label{fig:examples}
\end{figure}

@@

In the HMM (\figureref{examples-hmm}),
$h_2$ is a bottleneck, so we can recover $O \eqdef \Pr(x_2 \mid h_2)$.
%the parameters
%are $O \eqdef \Pr(x_i|h_i)$ and $T \eqdef \Pr(h_{i+1} | h_i)$
%for all $i$.
While the first hidden variable $h_1$
is not a bottleneck,
it still has an exclusive view $x_1$ with respect to the clique $\{h_1,h_2\}$,
assuming parameter sharing across emissions ($\Pr(x_1 \mid h_1) = O$).

@@

In the latent tree model (\figureref{examples-tree}),
%the parameters
%are $\pi \eqdef \Pr(h_i)$, $T \eqdef \Pr(h_i | h_1)$ for $i \in \{2,3,4\}$,
%and $O \eqdef \Pr(x^a_i | h_i) = \Pr(x^b_i | h_i)$ for $i \in \{2,3,4\}$.
$h_1$ is not directly connected to an observed variable,
but it is still a bottleneck, with views $x^a_2, x^a_3, x^a_4$, for example.
The clique $\{h_1,h_2\}$ has exclusive views $\{x_2^a,x_3^a\}$.
%We can recover $T$ from the clique $\{h_1, h_2\}$ by using views $x^a_2$
  %(exclusive to $h_2$) and $x^a_3$ (exclusive to $h_1$).

\paragraph{Non-example}
\label{sec:non-example}

%There are certainly models which are identifiable but do not have exclusive views.
In \figureref{examples-noisy-or},
  $h_1$ does not have exclusive views.
  Without parameter sharing, the techniques in this paper
  are insufficient.
In the special case where the graphical model represents a binary-valued noisy-or network,
  we can use the algorithm of \citet{halpern13noisyor},
  which first learns $h_2$ and subtracts off its influence,
  thereby making $h_1$ a bottleneck.
